\newcommand{\eclipseURL}{http://www.eclipse.org}
\newcommand{\antURL}{http://ant.apache.org}

\xname{eclipse_overview}
\subsection{Overview}
\index{Eclipse IDE, running under Jikes RVM}

Since version 2.2.1, Jikes RVM has run
\xlink{Eclipse}{\eclipseURL}.  The current (October 2004) \jrvm{} CVS
head, and \jrvm{} versions 2.3.4 and later,
run Eclipse 3.0.  This is really a technology preview and
for that reason we start with some caveats, the most obvious being
that production-mode use of Eclipse with Jikes RVM
is not recommended at this time.  Furthermore:
\begin{itemize}

\item If running Eclipse 3.0, then you need to tell Jikes RVM to use
the Xalan XML processor; we describe this process below.

\item We would suggest using either the {\tt prototype} or 
{\tt development} configurations to run Eclipse.

\item Under Linux, use the Eclipse builds for the GTK windowing
toolkit.  There are Linux builds for the Motif toolkit, but we never
test these and work on Eclipse seems to focus on the GTK version.

\item We recommend Eclipse 3.0.1, currently the latest release
from \xlink{{\tt \eclipseURL}}{\eclipseURL}; this is what we use and
test at Watson.  In the past we have successfully run Eclipse 2.1.0,
2.1.1, and the Eclipse 3.0 prerelease CVS head; the most testing has
been done on Eclipse 2.1.1.  The Eclipse 2.0.x
releases may also work.  Do not try to use Eclipse 1.x versions.

\item The Eclipse debugging support for Java programs --- which relies
on talking \xlink{JDWP}{\JDWPURL} to the VM --- will not work because
Jikes RVM has yet to implement this protocol.  (\link{It's on our TODO
  list, though}[ --- see section~\Ref, page~\Pageref{}]{JDWP}.)

\item Running Eclipse requires that you build Jikes RVM with
\varName{RVM\_\-FOR\_\-SIN\-GLE\_\-VIR\-TU\-AL\_\-PRO\-CES\-SOR} equal to 0. \xlink{The
select interception mechanism (which is required to run Eclipse on
Jikes RVM) does not work if this variable is set to 1 }[(see defect \#
3583)]{\RVMBugTrackerURL/?func=detail\&aid=1147482\&group_id=128805\&atid=712768}.
%{http://www-124.ibm.com/developerworks/bugs/?func=detailbug\&bug\_id=3583\&group\_id=95}.

\end{itemize}

{\it Given} these caveats, Eclipse has been running on Jikes RVM at
Watson in various configurations since the summer of 2002, and the
standard Eclipse functionality appears to work (The single biggest
exception is debugging support, as mentioned above).  We have
successfully developed and run Java programs, checked projects out of
CVS repositories, used the refactoring support, run the Web-based help
system, applied updates to Eclipse, and so on.  Furthermore,
performance is not bad, at least when adaptive builds of Jikes RVM,
such as {\tt production} or {\tt development}, are used.  Most
importantly, we very much want users to try running Eclipse on Jikes
RVM: we eagerly solicit feedback regarding bugs and would be
especially grateful for any contributions that enhance the usability
of Eclipse on Jikes RVM.\@

Setting up Eclipse and Jikes RVM is relatively straightforward; you
have to install Jikes RVM and Eclipse themselves first, and, after
that, there are just a few steps:
\begin{enumerate}
\index{Ant, a build tool}
\item \xlink{Install Xalan-Java.}[  You can get Xalan-Java from {\tt
\xalanURL}.]{\xalanURL}  We suggest you download one of the {\it binary}
distributions; otherwise, if you build Xalan-Java from source,  you'll have to
separately download and build Xerces.  \link{There is more discussion
of Xalan-Java in the installation section}[ --- see section~\Ref, page~\Pageref{}]{xalan-why}
\item \xlink{Install Ant.}[  You can get Ant from {\tt \antURL}.]{\antURL} 
\item Add three variables to your Jikes RVM configuration file:
 \begin{description}
 \item[\IndexTexttt{XALAN\_J\_DIR}] is the root of your Xalan binary
 installation.  
 \item[\IndexTexttt{ANT\_CMD}] is the full pathname of the executable Ant command
 \item[\IndexTexttt{ECLIPSE\_INSTALL\_DIR}] is the root of your Eclipse installation
 \end{description}
\item In a fresh build directory, run {\tt ./jbuild.plugin} to install
support for Eclipse to use Jikes RVM.\@
\end{enumerate}

\xname{rvmeclipse}
\subsection{The {\tt rvmeclipse} command}

\Indextt{rvmeclipse}%
Once you have done this, you will be able to run Eclipse on Jikes RVM
using the {\tt \$RVM\_\-ROOT/\-rvm/\-bin/\-rvmeclipse} command.  This wrapper
calls the normal Eclipse command, instructing it to use Jikes RVM to
run Eclipse.  Set \varName{RVM\_\-ROOT} and \varName{RVM\_BUILD} before you invoke {\tt
rvmeclipse}.

When you run {\tt rvmeclipse}, if you don't already
have an Eclipse workspace created, you'll see a message like:
\begin{example}
{\tt JikesRVM: exit 23}
\end{example}
This message is normal; just ignore it.  Jikes RVM always prints out
the exit status when it exits with a non-zero exit value; other Java\TMweb{}
Virtual Machines don't.

\xname{plugin}
\subsection{What the Plugin Does}

To understand the plugin, you need to understand that when you're
using Eclipse as a Java IDE, Eclipse uses two Java Virtual Machines
(JVMs): one (the {\em primary JVM}) to run Eclipse itself, and
another, the {\em inferior JVM}, that the Eclipse Java Development
Tool subsystem uses to run your Java applications.

The {\tt rvmeclipse} command launches the Eclipse IDE itself on top of
Jikes RVM.\@  So, {\tt rvmeclipse} lets Eclipse use Jikes RVM as the
{\em primary JVM}.  The Jikes RVM Eclipse plugin lets you use Jikes RVM as
your {\em inferior JVM} for running your user programs {\em under} Eclipse.

The plugin will only work in an Eclipse session that itself is already
using Jikes RVM as the {\em primary JVM}.  The easiest way to do this
is to start Eclipse with the command {\tt rvmeclipse}, as discussed
above.

In short, the plugin is irrelevant to running Eclipse itself under
Jikes RVM.\@  It only affects how Eclipse runs inferior Java processes;
if you're not using Eclipse as an IDE for Java programming, then
the plugin is irrelevant to your needs.

If you want to continue developing your application Java programs
under another JVM, such as the Blackdown or IBM JDK, then you also
won't have a use for the plugin, since the Eclipse Java IDE already
knows how to use those JVMs to run your programs.

