\subsection{What's New in this Release?}

Below is a list of major changes from the previous release
\begin{description}
\item [Linux/IA32 minimal functionality]
%% Tony/Maria
This base release includes a partial release of the Linux/IA32 support.
It includes the baseline compiler, threading, locking, copying
and noncopying garbage collectors, and the debugger (jdp).

This release does NOT include: Reflection, SMP support, Java\trademark native
Interface(JNI), 
the remaining garbage collectors,
and the optimizing compiler. 

\item [Adaptive optimization system]
This release includes the adaptive optimization system as described
in the \xlink{2000 OOPSLA paper}
{http://www.research.ibm.com/jalapeno/publication.html\#oopsla00\_aos}
\T~\cite{jalapeno-adaptive-00}

\item [Opt compiler counter instrumentation infrastructure:]
%% Matt
Basic infrastructure to simplify the process of inserting
instrumented counters in opt-compiled code. See
Section~\ref{counting_events}.

\item [Hybrid garbage collector]
%% Steve Smith
A generational collector that combines a Nursery region for object
allocation, managed using the logic of the copying generational collector,
and a Mature Object Heap for objects that survive one or more
collections, managed using the logic of the non-copying
mark-sweep collector.

\item [Elimination of specific Jikes\trademark dependency]
%% Igor
As of this release, \jp\ is not dependent on a particular version of
Jikes\trademark.  It will work with any Jikes\trademark binary.  \jp\ still requires that
its source-to-bytecode compiler be Jikes\trademark (as opposed to javac), because
it uses features that are only available in Jikes\trademark.

\item [VM callbacks]
%% Igor
A mechanism to allow arbitrary code to be executed on certain VM events.
See section~\ref{sssec:callbacks}.

\item [Updated libraries]
%% Maria  source?

\end{description}
